\documentclass{amsart}

\usepackage{amsmath,amssymb,graphicx,stmaryrd}

\usepackage{./boldfonts}
\usepackage{./macro}


\begin{document}

\title[The Laplace-Beltrami Operator]{Finite Element Methods for the \\ Laplace-Beltrami Operator}%\label{chap:1}
\author{Andrea Bonito}
\address{Department of Mathematics, Texas A\&M University, College Station, TX 77843, USA; bonito@math.tamu.edu}%
\thanks{AB is partially supported by NSF grant DMS-1817691.}
\author{Alan Demlow}%
\address{Department of Mathematics, Texas A\&M University, College Station, TX 77843, USA; demlow@math.tamu.edu}%
\thanks{AD is partially supported by NSF grant DMS-1720369.}
\author{Ricardo H. Nochetto}%
\address {University of Maryland, College Park, MD 20742, USA; rhn@math.umd.edu}%
\thanks{RHN is partially supported by NSF grant DMS--1411808.}



\begin{abstract}
Partial differential equations posed on surfaces arise in a number of applications.  In this survey we describe three popular finite element methods for approximating solutions to the Laplace-Beltrami problem posed on an $n$-dimensional surface $\gamma$ embedded in $\mathbb{R}^{n+1}$:  the {\it parametric}, {\it trace}, and {\it narrow band} methods.  The parametric method entails constructing an approximating polyhedral surface $\Gamma$ whose faces comprise the finite element triangulation.  The finite element method is then posed over the approximate surface $\Gamma$ in a manner very similar to standard FEM on Euclidean domains.  In the trace method it is assumed that the given surface $\gamma$ is embedded in an $n+1$-dimensional domain $\Omega$ which has itself been triangulated.    An $n$-dimensional approximate surface $\Gamma$ is then constructed roughly speaking by interpolating $\gamma$ over the triangulation of $\Omega$, and the finite element space over $\Gamma$ consists of the trace (restriction) of a standard finite element space on $\Omega$ to $\Gamma$.  In the narrow band method the PDE posed on the surface is extended to a triangulated $n+1$-dimensional band about $\gamma$ whose width is proportional to the diameter of elements in the triangulation.  In all cases we provide optimal a priori error estimates for the lowest-order finite element methods, and we also present a posteriori error estimates for the parametric and trace methods.  Our presentation focuses especially on the relationship between the regularity of the surface $\gamma$, which is never assumed better than of class $C^2$, the manner in which $\gamma$ is represented in theory and practice, and the properties of the resulting methods.  
\end{abstract}

\keywords{
Surface partial differential equations, Laplace-Beltrami operator, surface finite element methods, parametric finite element methods, trace finite element methods, narrow band finite element methods, a-priori and a-posteriori error estimates.
}

\date{\today}

\maketitle

% \minitoc


\newpage 


\begingroup
\parindent=0em
\etocsettocstyle{\rule{\linewidth}{\tocrulewidth}\vskip0.5\baselineskip}{\rule{\linewidth}{\tocrulewidth}}
\localtableofcontents 
\endgroup


%
\input{introduction_v2.tex}
\input{preliminaries_v6.tex}
\input{parametric_v9.tex}
\input{trace_v6.tex}
\input{narrow_v9.tex}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{alpha}
\newcommand{\etalchar}[1]{$^{#1}$}
\begin{thebibliography}{ADM{\etalchar{+}}15}

\bibitem[Ada75]{Ad75}
Robert~A. Adams.
\newblock {\em Sobolev spaces}.
\newblock Academic Press [A subsidiary of Harcourt Brace Jovanovich,
  Publishers], New York-London, 1975.
\newblock Pure and Applied Mathematics, Vol. 65.

\bibitem[ADM{\etalchar{+}}15]{ADMSSV15}
Paola~F. Antonietti, Andreas Dedner, Pravin Madhavan, Simone Stangalino,
  Bj\"{o}rn Stinner, and Marco Verani.
\newblock High order discontinuous {G}alerkin methods for elliptic problems on
  surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 53(2):1145--1171, 2015.

\bibitem[Aub82]{MR681859}
Thierry Aubin.
\newblock {\em Nonlinear analysis on manifolds. {M}onge-{A}mp\`ere equations},
  volume 252 of {\em Grundlehren der Mathematischen Wissenschaften [Fundamental
  Principles of Mathematical Sciences]}.
\newblock Springer-Verlag, New York, 1982.

\bibitem[BCM{\etalchar{+}}16]{BCMMN16}
Andrea Bonito, J.~Manuel Casc{\'o}n, Khamron Mekchay, Pedro Morin, and
  Ricardo~H. Nochetto.
\newblock High-order {AFEM} for the {L}aplace-{B}eltrami operator: Convergence
  rates.
\newblock {\em Found. Comput. Math.}, 16(6):1473--1539, 2016.

\bibitem[BCMN13]{BCMN:Magenes}
A.~Bonito, J.~M. Casc{\'o}n, P.~Morin, and R.~H. Nochetto.
\newblock A{FEM} for geometric {PDE}: the {L}aplace-{B}eltrami operator.
\newblock In {\em Analysis and Numerics of Partial Differential Equations},
  volume~4 of {\em Springer INdAM Ser.}, pages 257--306. Springer, Milan, 2013.

\bibitem[BCOS01]{MR1868103}
Marcelo Bertalm\'{i}o, Li-Tien Cheng, Stanley Osher, and Guillermo Sapiro.
\newblock Variational problems and partial differential equations on implicit
  surfaces.
\newblock {\em J. Comput. Phys.}, 174(2):759--780, 2001.

\bibitem[BD19]{BD:18}
A.~Bonito and A.~Demlow.
\newblock A posteriori error estimates for the laplace-beltrami operator on
  parametric {${C}^2$} surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 57(3):973--996, 2019.

\bibitem[BDO]{BDO19}
A.~Bonito, A.~Demlow, and J.~Owen.
\newblock Convergence and optimality for higher order afem for eigenvalue
  clusters of the {L}aplace-{B}eltrami operator.
\newblock in preparation.

\bibitem[Ben84]{Ben84}
A.~Bendali.
\newblock Numerical analysis of the exterior boundary value problem for the
  time-harmonic {M}axwell equations by a boundary finite element method. {II}.
  {T}he discrete problem.
\newblock {\em Math. Comp.}, 43(167):47--68, 1984.

\bibitem[BHL15]{BHL15}
Erik Burman, Peter Hansbo, and Mats~G. Larson.
\newblock A stabilized cut finite element method for partial differential
  equations on surfaces: the {L}aplace-{B}eltrami operator.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 285:188--207, 2015.

\bibitem[BHL{\etalchar{+}}16]{BHLMZ16}
Erik Burman, Peter Hansbo, Mats~G. Larson, Andr\'{e} Massing, and Sara Zahedi.
\newblock Full gradient stabilized cut finite element methods for surface
  partial differential equations.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 310:278--296, 2016.

\bibitem[BHLM17]{BHLM17}
Erik Burman, Peter Hansbo, Mats~G. Larson, and Andr\'{e} Massing.
\newblock A cut discontinuous {G}alerkin method for the {L}aplace-{B}eltrami
  operator.
\newblock {\em IMA J. Numer. Anal.}, 37(1):138--169, 2017.

\bibitem[BP12]{BP:11}
A.~Bonito and J.E. Pasciak.
\newblock Convergence analysis of variational and non-variational multigrid
  algorithm for the {L}aplace-{B}eltrami operator.
\newblock {\em Math. Comp.}, 81:1263--1288, 2012.

\bibitem[Bur09]{MR2485787}
Martin Burger.
\newblock Finite element approximation of elliptic partial differential
  equations on implicit surfaces.
\newblock {\em Comput. Vis. Sci.}, 12(3):87--100, 2009.

\bibitem[CD15]{CD15}
Fernando Camacho and Alan Demlow.
\newblock {$L_2$} and pointwise a posteriori error estimates for {FEM} for
  elliptic {PDE}s on surfaces.
\newblock {\em IMA J. Numer. Anal.}, 35(3):1199--1227, 2015.

\bibitem[CD16]{CD16}
Bernardo Cockburn and Alan Demlow.
\newblock Hybridizable discontinuous {G}alerkin and mixed finite element
  methods for elliptic problems on surfaces.
\newblock {\em Math. Comp.}, 85(302):2609--2638, 2016.

\bibitem[CO15]{CO15}
Alexey~Y. Chernyshenko and Maxim~A. Olshanskii.
\newblock An adaptive octree finite element method for {PDE}s posed on
  surfaces.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 291:146--172, 2015.

\bibitem[DD07]{DemlowDziuk:07}
A.~Demlow and G.~Dziuk.
\newblock An adaptive finite element method for the {L}aplace-{B}eltrami
  operator on implicitly defined surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 45(1):421--442 (electronic), 2007.

\bibitem[DDEH10]{MR2608464}
Klaus Deckelnick, Gerhard Dziuk, Charles~M. Elliott, and Claus-Justus Heine.
\newblock An {$h$}-narrow band finite-element method for elliptic equations on
  implicit surfaces.
\newblock {\em IMA J. Numer. Anal.}, 30(2):351--376, 2010.

\bibitem[DE13]{DE13}
Gerhard Dziuk and Charles~M. Elliott.
\newblock Finite element methods for surface {PDE}s.
\newblock {\em Acta Numer.}, 22:289--396, 2013.

\bibitem[Dem09]{De09}
Alan Demlow.
\newblock Higher-order finite element methods and pointwise error estimates for
  elliptic problems on surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 47(2):805--827, 2009.

\bibitem[DER14]{MR3249369}
Klaus Deckelnick, Charles~M. Elliott, and Thomas Ranner.
\newblock Unfitted finite element methods using bulk meshes for surface partial
  differential equations.
\newblock {\em SIAM J. Numer. Anal.}, 52(4):2137--2162, 2014.

\bibitem[DM16]{DM16}
Andreas Dedner and Pravin Madhavan.
\newblock Adaptive discontinuous {G}alerkin methods on surfaces.
\newblock {\em Numer. Math.}, 132(2):369--398, 2016.

\bibitem[DMS13]{DMS13}
Andreas Dedner, Pravin Madhavan, and Bj{\"o}rn Stinner.
\newblock Analysis of the discontinuous {G}alerkin method for elliptic problems
  on surfaces.
\newblock {\em IMA J. Numer. Anal.}, 33(3):952--973, 2013.

\bibitem[DO12]{DO12}
Alan Demlow and Maxim Olshanskii.
\newblock An adaptive surface finite element method based on volume meshes.
\newblock {\em SIAM J. Numer. Anal.}, 50:1624--1647, 2012.

\bibitem[Dzi88]{Dz88}
Gerhard Dziuk.
\newblock Finite elements for the {B}eltrami operator on arbitrary surfaces.
\newblock In {\em Partial Differential Equations and Calculus of Variations},
  volume 1357 of {\em Lecture Notes in Math.}, pages 142--155. Springer,
  Berlin, 1988.

\bibitem[Eva98]{Ev98}
Lawrence~C. Evans.
\newblock {\em Partial Differential Equations}, volume~19 of {\em Graduate
  Studies in Mathematics}.
\newblock American Mathematical Society, Providence, RI, 1998.

\bibitem[Fed59]{Fed59}
Herbert Federer.
\newblock Curvature measures.
\newblock {\em Trans. Amer. Math. Soc.}, 93:418--491, 1959.

\bibitem[FFF16]{FFF16}
Alberto Ferroni, Luca Formaggia, and Alessio Fumagalli.
\newblock Numerical analysis of {D}arcy problem on surfaces.
\newblock {\em ESAIM Math. Model. Numer. Anal.}, 50(6):1615--1630, 2016.

\bibitem[GLR18]{GLR18}
J\"{o}rg Grande, Christoph Lehrenfeld, and Arnold Reusken.
\newblock Analysis of a high-order trace finite element method for {PDE}s on
  level set surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 56(1):228--255, 2018.

\bibitem[GR16]{GR16}
J\"{o}rg Grande and Arnold Reusken.
\newblock A higher order finite element method for partial differential
  equations on surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 54(1):388--414, 2016.

\bibitem[Gra17]{Gr17}
J\"org Grande.
\newblock Analysis of highly accurate finite element based algorithms for
  computing distances to level sets.
\newblock {\em SIAM J. Numer. Anal.}, 55(1):376--399, 2017.

\bibitem[Gri85]{G85}
Pierre Grisvard.
\newblock {\em Elliptic Problems in Nonsmooth Domains}.
\newblock Pitman Publishing Inc., Marshfield, MA, 1985.

\bibitem[GT98]{GT98}
David Gilbarg and Neil~S. Trudinger.
\newblock {\em Elliptic Partial Differential Equations of Second Order}.
\newblock Springer-Verlag, Berlin, 2nd edition, 1998.

\bibitem[HH02]{HH02}
Anita Hansbo and Peter Hansbo.
\newblock An unfitted finite element method, based on {N}itsche's method, for
  elliptic interface problems.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 191(47-48):5537--5552,
  2002.

\bibitem[HH04]{HH04}
Anita Hansbo and Peter Hansbo.
\newblock A finite element method for the simulation of strong and weak
  discontinuities in solid mechanics.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 193(33-35):3523--3540,
  2004.

\bibitem[HS12]{HoSt12}
Michael Holst and Ari Stern.
\newblock Geometric variational crimes: {H}ilbert complexes, finite element
  exterior calculus, and problems on hypersurfaces.
\newblock {\em Found. Comput. Math.}, 12(3):263--293, 2012.

\bibitem[JK95]{JK95}
David Jerison and Carlos~E. Kenig.
\newblock The inhomogeneous {D}irichlet problem in {L}ipschitz domains.
\newblock {\em J. Funct. Anal.}, 130(1):161--219, 1995.

\bibitem[Luc57]{Luc57}
Kenneth~R. Lucas.
\newblock Submanifolds of dimension $n-1$ in $\mathbb{R}^n$ with normals
  satisfying a lipschitz condition, studies in eigenvalue problems.
\newblock Technical report, University of Kansas, 1957.

\bibitem[N{\'e}d76]{Ne76}
J.-C. N{\'e}d{\'e}lec.
\newblock Curved finite element methods for the solution of singular integral
  equations on surfaces in {$R\sp{3}$}.
\newblock {\em Comput. Methods Appl. Mech. Engrg.}, 8(1):61--80, 1976.

\bibitem[OR10]{OR10}
Maxim~A. Olshanskii and Arnold Reusken.
\newblock A finite element method for surface {PDE}s: matrix properties.
\newblock {\em Numer. Math.}, 114(3):491--520, 2010.

\bibitem[OR17]{OR17}
Maxim~A. Olshanskii and Arnold Reusken.
\newblock Trace finite element methods for {PDE}s on surfaces.
\newblock In {\em Geometrically Unfitted Finite Element Methods and
  Applications}, volume 121 of {\em Lect. Notes Comput. Sci. Eng.}, pages
  211--258. Springer, Cham, 2017.

\bibitem[ORG09]{ORG09}
Maxim~A. Olshanskii, Arnold Reusken, and J\"{o}rg Grande.
\newblock A finite element method for elliptic equations on surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 47(5):3339--3358, 2009.

\bibitem[ORX12]{ORX12}
Maxim~A. Olshanskii, Arnold Reusken, and Xianmin Xu.
\newblock On surface meshes induced by level set functions.
\newblock {\em Comput. Vis. Sci.}, 15(2):53--60, 2012.

\bibitem[OS16]{MR3471100}
Maxim~A. Olshanskii and Danil Safin.
\newblock A narrow-band unfitted finite element method for elliptic {PDE}s
  posed on surfaces.
\newblock {\em Math. Comp.}, 85(300):1549--1570, 2016.

\bibitem[Pee66]{MR0221282}
Jaak Peetre.
\newblock Espaces d'interpolation et th\'{e}or\`eme de {S}oboleff.
\newblock {\em Ann. Inst. Fourier (Grenoble)}, 16(fasc. 1):279--317, 1966.

\bibitem[Reu15]{Re15}
Arnold Reusken.
\newblock Analysis of trace finite element methods for surface partial
  differential equations.
\newblock {\em IMA J. Numer. Anal.}, 35(4):1568--1590, 2015.

\bibitem[RM08]{RM08}
Steven~J. Ruuth and Barry Merriman.
\newblock A simple embedding method for solving partial differential equations
  on surfaces.
\newblock {\em J. Comput. Phys.}, 227(3):1943--1961, 2008.

\bibitem[Ste70]{Stein:70}
Elias~M. Stein.
\newblock {\em Singular Integrals and Differentiability Properties of
  Functions}.
\newblock Princeton Mathematical Series, No. 30. Princeton University Press,
  Princeton, N.J., 1970.

\bibitem[Tar78]{MR532371}
Luc Tartar.
\newblock {\em Topics in Nonlinear Analysis}, volume~13 of {\em Publications
  Math\'{e}matiques d'Orsay 78}.
\newblock Universit\'{e} de Paris-Sud, D\'{e}partement de Math\'{e}matique,
  Orsay, 1978.

\bibitem[Vee16]{Vee15}
Andreas Veeser.
\newblock Approximating gradients with continuous piecewise polynomial
  functions.
\newblock {\em Found. Comput. Math.}, 16(3):723--750, 2016.

\bibitem[Ver13]{MR3059294}
R\"{u}diger Verf\"{u}rth.
\newblock {\em A Posteriori Error Estimation Techniques for Finite Element
  Methods}.
\newblock Numerical Mathematics and Scientific Computation. Oxford University
  Press, Oxford, 2013.

\bibitem[WCH10]{WCH10}
Huayi Wei, Long Chen, and Yunqing Huang.
\newblock Superconvergence and gradient recovery of linear finite elements for
  the {L}aplace-{B}eltrami operator on general surfaces.
\newblock {\em SIAM J. Numer. Anal.}, 48(5):1920--1943, 2010.

\end{thebibliography}







\end{document}
